---
title: "Midterm"
output: html_document
---

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(pscl)
library(plotROC)
library(pROC)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tigris)
library(viridis)
library(gridExtra)
library(dplyr)
library(raster)
library(spdep)
library(FNN)
library(grid)
library(mapview)
library(tidycensus)
library(ggplot2)

mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.75),
  axis.ticks=element_blank())

palette2 <- c("#981FAC","#FF006A")
```

## Loading and Cleaning Data
### Calgary Data
```{r calgary initial data}
calgaryboundary <- st_read("FE_calgshp/calg_boundary.shp") %>% st_transform(crs = 3400)  #city boundary 
cal_streams <- st_read("FE_calgshp/calg_streams.shp") %>% st_transform(crs = 3400) #streams
cal_avgel <- st_read("FE_calgshp/calg_fishnet_avgel.shp") %>% st_transform(crs = 3400) #avg elevation
cal_fac <- st_read("FE_calgshp/calg_fishnet_fac.shp") %>% st_transform(crs = 3400) #flow accumulation
cal_impervious <- st_read("FE_calgshp/calg_fishnet_imperv.shp") %>% st_transform(crs = 3400)
cal_inundation <- st_read("FE_calgshp/calg_fishnet_inun.shp") %>% st_transform(crs = 3400)#inundation
cal_slope <- st_read("FE_calgshp/calg_fishnet_slope.shp") %>% st_transform(crs = 3400) #slope in degrees
cal_strdist <- st_read("FE_calgshp/calg_fishnet_strdist.shp") %>% st_transform(crs = 3400) #distance from stream
```

```{r calgary cleaned data}
#creating dataframe
AREA <- c("")
MAJORITY <- c("")
avgel <- c("") #average elevation
facMean <- c("") #mean column of cal_fac
impervSum <- c("") #sum column of cal_impervious
slopeMean <- c("") #max column of cal_slope
strDist <- c("") #zsclagst_4 of cal_strdist
geometry <- c("")
calgary <- data.frame (AREA, MAJORITY, avgel, facMean, impervSum, slopeMean, strDist, geometry)

geometry <- cal_inundation$geometry

#create new inundation without geometry to populate some variables
cal_inundation1 <- cal_inundation %>% st_drop_geometry() 
AREA <- cal_inundation1[, c("zscalgin_3")]
MAJORITY <- cal_inundation1[, c("zscalgin_4")]

#create new feature engineered datasets without geometry to populate some variables
cal_avgel1 <- cal_avgel %>% st_drop_geometry()
avgel <- cal_avgel1[, c("zscalgav_4")] #mean

cal_fac1 <- cal_fac %>% st_drop_geometry()
facMean <- cal_fac1[, c("zscalgfa_3")] #mean

cal_impervious1 <- cal_impervious %>% st_drop_geometry()
impervSum <- cal_impervious1[, c("zscalgim_4")] #sum

cal_slope1 <- cal_slope %>% st_drop_geometry()
slopeMean <- cal_slope1[, c("zscalgsl_4")] #mean

cal_strdist1 <- cal_strdist %>% st_drop_geometry()
strDist <- cal_strdist1[, c("zscalgst_4")] #mean

calgary <- data.frame (AREA, MAJORITY, avgel, facMean, impervSum, slopeMean, strDist, geometry) %>% st_as_sf() %>% st_transform(crs = 3400)
```

### Portland-Vancouver Data
```{r portland initial data}
plvcboundary <- st_read("FE_plvcshp/PLVC_bounds.shp") %>% st_transform(crs = 8756)  #city boundary 
plvc_fishnet <- st_read("FE_plvcshp/plvc_fishnet_clear.shp") %>% st_transform(crs = 8756) #streams
plvc_streams <- st_read("FE_plvcshp/PLVC_streams.shp") %>% st_transform(crs = 8756) #streams
plvc_avgel <- st_read("FE_plvcshp/PLVC_fishnet_avgel.shp") %>% st_transform(crs = 8756) #avg elevation
plvc_fac <- st_read("FE_plvcshp/plvc_fishnet_fac.shp") %>% st_transform(crs = 8756) #flow accumulation
plvc_impervious <- st_read("FE_plvcshp/plvc_fishnet_imperv.shp") %>% st_transform(crs = 8756)
plvc_slope <- st_read("FE_plvcshp/plvc_fishnet_slope.shp") %>% st_transform(crs = 8756) #slope in degrees
plvc_strdist <- st_read("FE_plvcshp/plvc_fishnet_strdist.shp") %>% st_transform(crs = 8756) #distance from stream
```

```{r portland cleaned data}
#creating dataframe
p_AREA <- c("")
p_avgel <- c("") #average elevation
p_facMean <- c("") #mean column of cal_fac
p_impervSum <- c("") #sum column of cal_impervious
p_slopeMean <- c("") #max column of cal_slope
p_strDist <- c("") #zsclagst_4 of cal_strdist
p_geometry <- c("")
plvc <- data.frame (p_AREA, p_avgel, p_facMean, p_impervSum, p_slopeMean, p_strDist, p_geometry)

p_geometry <- plvc_fishnet$geometry

#create new inundation without geometry to populate some variables
p_fishnet1 <- plvc_fishnet %>% st_drop_geometry() 
p_AREA <- p_fishnet1[, c("Area")]

#create new feature engineered datasets without geometry to populate some variables
plvc_avgel1 <- plvc_avgel %>% st_drop_geometry()
p_avgel <- plvc_avgel1[, c("zsplvcav_4")] #mean

plvc_fac1 <- plvc_fac %>% st_drop_geometry()
p_facMean <- plvc_fac1[, c("zsplvcfa_3")] #mean

plvc_impervious1 <- plvc_impervious %>% st_drop_geometry()
p_impervSum <- plvc_impervious1[, c("zsplvcim_4")] #sum

plvc_slope1 <- plvc_slope %>% st_drop_geometry()
p_slopeMean <- plvc_slope1[, c("zsplvcsl_4")] #mean

plvc_strdist1 <- plvc_strdist %>% st_drop_geometry()
p_strDist <- plvc_strdist1[, c("zsplvcst_4")] #mean

plvc <- data.frame (p_AREA, p_avgel, p_facMean, p_impervSum, p_slopeMean, p_strDist, p_geometry) %>% st_as_sf() %>% st_transform(crs = 8756)

plvc <- plvc %>%
  rename(AREA = p_AREA)
plvc <- plvc %>%
  rename(avgel = p_avgel)
plvc <- plvc %>%
  rename(facMean = p_facMean)
plvc <- plvc %>%
  rename(impervSum = p_impervSum)
plvc <- plvc %>%
  rename(slopeMean = p_slopeMean)
plvc <- plvc %>%
  rename(strDist = p_strDist)
```

### Feature Engineering
```{r}
calgary <- calgary %>%
  mutate(elevation = case_when(
    avgel <= 1077.99999 ~ "low",
    avgel >= 1078 & avgel <= 1189.99999 ~ "average",
    TRUE ~ "high"))

calgary <- calgary %>%
  mutate(flow = case_when(
    facMean <= 33016.99999 ~ "low",
    facMean >= 33017 & facMean <=66030.99999 ~ "average",
    TRUE ~ "high"))

calgary <- calgary %>%
  mutate(distance = case_when(
    strDist <= 1356.99999 ~ "nearest",
    strDist >= 1357 & strDist <=2673.99999 ~ "near",
    strDist >= 2674 & strDist <=3990.99999 ~ "average",
    strDist >= 3991 & strDist <=5307.99999 ~ "far",
    TRUE ~ "furthest"))

calgary <- calgary %>%
  mutate(slope = case_when(
    slopeMean < 3 ~ "flat",
    TRUE ~ "steep"))

calgary <- calgary %>%
  mutate(pervious = case_when(
    impervSum <= 5 ~ "yes",
    TRUE ~ "no"))

plvc <- plvc %>%
  mutate(elevation = case_when(
    avgel <= 117.99999 ~ "low",
    avgel >= 118 & avgel <= 226.99999 ~ "average",
    avgel >=227 ~ "high"))

plvc <- plvc %>%
  mutate(flow = case_when(
    facMean <= 24519.99999 ~ "low",
    facMean >= 24520 & facMean <=49031.999 ~ "average",
    TRUE ~ "high"))

plvc <- plvc %>%
  mutate(distance = case_when(
    strDist <= 774.99999 ~ "nearest",
    strDist >= 775 & strDist <=1517.99999 ~ "near",
    strDist >= 1518 & strDist <=2260.99999 ~ "average",
    strDist >= 2261 & strDist <=3003.99999 ~ "far",
    TRUE ~ "furthest"))

plvc <- plvc %>%
  mutate(slope = case_when(
    slopeMean < 3 ~ "flat",
    TRUE ~ "steep"))

plvc <- plvc %>%
  mutate(pervious = case_when(
    impervSum <= 5 ~ "yes",
    TRUE ~ "no"))
```

# Data Visualization
## Maps
```{r calgary ggplot, fig.width=5, fig.height=5}
ggplot() + 
  geom_sf(data=calgary, fill = "dark blue", color = "dark blue") +
  geom_sf(data = calgary, aes(fill=as.factor(MAJORITY)), color = "light blue")+
  scale_fill_manual(values = c("white", "light blue"),
                    labels = c("No Flooding","Flooding"),
                    name = "") +
  labs(title="Observed Flooding in Calgary (Fishnet)") +
  mapTheme

grid.arrange(ncol = 3,
ggplot() +
  geom_sf(data = calgary) +
  geom_sf(data = cal_avgel,
          aes(fill=zscalgav_4), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans="reverse",
                     name = "Mean") +
  labs(title = "Average \n Elevation  ") +
  mapTheme,

ggplot() +
  geom_sf(data = calgary) +
  geom_sf(data = cal_fac,
          aes(fill=zscalgfa_3), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Flow \n Accumulation") +
  mapTheme,

ggplot() +
  geom_sf(data = calgary) +
  geom_sf(data = cal_slope,
          aes(fill=zscalgsl_4), 
          colour=NA,) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Slope Degrees") +
  mapTheme,

ggplot() +
  geom_sf(data = calgary) +
  geom_sf(data = cal_strdist,
          aes(fill=zscalgst_4), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Distance From \n Stream") +
  mapTheme,

ggplot() +
  geom_sf(data = calgary) +
  geom_sf(data = cal_impervious,
          aes(fill=zscalgim_4), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Sum") +
  labs(title = "Impervious \n Surface") +
  mapTheme)

```

```{r plvc ggplot, fig.width=5, fig.height=5}
grid.arrange(ncol = 3,
ggplot() +
  geom_sf(data = plvc) +
  geom_sf(data = plvc_avgel,
          aes(fill=zsplvcav_4), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans="reverse",
                     name = "Mean") +
  labs(title = "Average \n Elevation") +
  mapTheme,

ggplot() +
  geom_sf(data = plvc) +
  geom_sf(data = plvc_fac,
          aes(fill=zsplvcfa_3), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Flow \n Accumulation") +
  mapTheme,

ggplot() +
  geom_sf(data = plvc) +
  geom_sf(data = plvc_slope,
          aes(fill=zsplvcsl_4), 
          colour=NA,) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Slope Degrees") +
  mapTheme,

ggplot() +
  geom_sf(data = plvc) +
  geom_sf(data = plvc_strdist,
          aes(fill=zsplvcst_4), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Distance \n From Stream") +
  mapTheme,

ggplot() +
  geom_sf(data = plvc) +
  geom_sf(data = plvc_impervious,
          aes(fill=zsplvcim_4), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans = "reverse",
                     name = "Mean") +
  labs(title = "Impervious \n Surface") +
  mapTheme)

```

## Plots (fix later) #fix FE
```{r exploratory analysis}
calgaryPlotVariables <- 
  calgary %>%
  as.data.frame() %>%
  dplyr::select(MAJORITY,facMean,avgel,slopeMean,strDist,impervSum) %>%
  gather(variable, value, -MAJORITY)

ggplot(calgaryPlotVariables %>%
         group_by(MAJORITY, variable) %>%
         summarize(mean = mean(value))) + 
     geom_bar(aes(as.factor(MAJORITY), 
                  mean, 
                  fill=as.factor(MAJORITY)),
              stat="identity") + 
     facet_wrap(~variable, ncol = 5, scales = "free") +
    labs(title = "Exploratory Analysis of Features", 
          caption="Without Feature Engineer") +
     scale_fill_manual(values = c("light blue", "dark blue"),
                      labels = c("Not Flooded","Flooded"),
                      name = "") +
    labs(x="Preserved", y="Value")


```

# Model building
To begin testing on the probability of flooding in Calgary, we partition the dataset into a 70/30 split to train and test on. 
```{r}
set.seed(3456)
trainIndex <- createDataPartition(calgary$MAJORITY, p = .70,
                                  list = FALSE,
                                  times = 1)
calgaryTrain <- calgary[ trainIndex,]
calgaryTest  <- calgary[-trainIndex,]
```

## Model binomial
This model incorporates many feature engineered variables such as how far is the distance from streams, the mean slope degrees, mean flow accumulation, whether a surface is an impervious or not, and elevation levels.
```{r}
options(scipen=999)
calgaryModel <- glm(MAJORITY ~ distance + slope + flow + pervious + elevation,
                    family="binomial"(link="logit"), data = calgaryTrain %>%
                      as.data.frame %>%
                      dplyr::select(-AREA, -geometry, -strDist, -facMean, -slopeMean, -impervSum, -avgel))

summary(calgaryModel)
```
The model found that the nearest and far values of distance from streams to be statistically significant. This is also true for the mean flow accumulation at both high and low levels. Pervious surfaces and low elevation were also found to be statistically significant. 

## Model Validation
The following histogram and plot indicates that the model may be overfit. While this can conclude a high level of accuracy, this comes at the expense of generalizability. The historgram of probabilities or predictions shows the variety in values that will be projected. 

HELP??? A negative or 0 value means that an area did not flood, while a positive or 1 value means that an area did flood. Strong models will have a peak closer to 0 for the negatives (no flooding), and a peak closer to 1 for the positives (flooding).
```{r histogram and testProbs}
classProbs <- predict(calgaryModel, calgaryTest, type="response")

hist(classProbs)

testProbs <- data.frame(obs = as.numeric(calgaryTest$MAJORITY),
                        pred = classProbs)

ggplot(testProbs, aes(x = pred, fill=as.factor(obs))) + 
  geom_density() +
  facet_grid(obs ~ .) + 
  xlab("Probability") + 
  geom_vline(xintercept = .5) +
  scale_fill_manual(values = c("light blue", "dark blue"),
                      labels = c("Not Flooded","Flooded"),
                      name = "")+
  plotTheme
```

### Confusion Matrix
Based off of the distribution above we change the threshold to 10 or 0.1 moving forward
```{r}
testProbs$predClass  = ifelse(testProbs$pred > .1 ,1,0)

caret::confusionMatrix(reference = as.factor(testProbs$obs), 
                       data = as.factor(testProbs$predClass), 
                       positive = "1")
```

### ROC Curve
AUC determines if a model is sufficient enough to run. 
```{r}
ggplot(testProbs, aes(d = obs, m = pred)) + 
  geom_roc(n.cuts = 50, labels = FALSE, color = 'blue') + 
  style_roc(theme = theme_grey) +
  geom_abline(slope = 1, intercept = 0, size = 1.5, color = 'grey') 

auc(testProbs$obs, testProbs$pred)

```

### Cross Validation
```{r}
ctrl <- trainControl(method = "cv", 
                     number = 100, 
                     savePredictions = TRUE)

cvFit <- train(as.factor(MAJORITY) ~ .,  data = calgary %>% 
                                                as.data.frame() %>%
                                                dplyr::select(-AREA, -geometry, -strDist, -facMean, -slopeMean, -impervSum, -avgel), 
               method="glm", family="binomial",
               trControl = ctrl)

cvFit
```

```{r}
ggplot(as.data.frame(cvFit$resample), aes(Accuracy)) + 
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x="Accuracy",
       y="Count")+
  plotTheme
```

# Predictions
```{r calgary predictions}
allPredictions <- 
  predict(cvFit, calgary, type="prob")[,2]
  
calgary <- 
  cbind(calgary,allPredictions) %>%
  mutate(allPredictions = round(allPredictions * 100))

calgary1 <- calgary %>%
  mutate(PredClass = ifelse(allPredictions > 10, 1, 0))

calgary1 <- calgary1 %>%
  mutate(Correct = ifelse(PredClass == MAJORITY, "1", "0"),
         Incorrect = ifelse(PredClass != MAJORITY, "1", "0"))

```

```{r plvc prediction}
plvc_allPredictions <- 
  predict(cvFit, plvc, type="prob")[,2] 

plvc <- 
  cbind(plvc,plvc_allPredictions) %>% 
  mutate(plvc_allPredictions = round(plvc_allPredictions * 100))
```

### Spatial Cross Validation
```{r}
cal_nhood <- st_read("nhood/geo_export_657fd344-2823-4047-abad-f7d6906b7c1a.shp") %>% st_transform(crs = 3400)
cal_nhood <- st_intersection(calgary1, cal_nhood)
cal_nhood <- cal_nhood %>% mutate(Correct = as.numeric(Correct),
                                  Incorrect = as.numeric(Incorrect))
cal_nhood <- cal_nhood %>% dplyr::select(name, sector, MAJORITY, allPredictions, PredClass, Correct, Incorrect, -geometry) %>% 
  group_by(name) %>% 
  summarise(meanPrediction = mean(allPredictions),
            accuracy = sum(Correct) / sum(Correct) + sum(Incorrect))

ggplot() +
  geom_sf(data = cal_nhood,
          aes(fill=accuracy), 
          colour=NA) +
  geom_sf(data=calgary  %>% 
               filter(MAJORITY == 1), 
               fill="gold",colour="NA") +
  scale_fill_viridis(option="mako", trans="reverse") +
  labs(title = "Average Probability of Flooding by Sector with Observed Flooding in Gold") +
  mapTheme

ggplot() +
  geom_sf(data = cal_nhood)+
  geom_sf(data = cal_nhood,
          aes(fill=accuracy), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans="reverse") +
  labs(title = "Accuracy of Flooding (Average by Neighborhood)") +
  mapTheme

ggplot() +
  geom_sf(data = cal_nhood)+
  geom_sf(data = cal_nhood,
          aes(fill=meanPrediction), 
          colour=NA) +
  scale_fill_viridis(option="mako", trans="reverse") +
  labs(title = "Average Probability of Flooding by Sector") +
  mapTheme
```

## Maps
Calgary and PLVC Prediction Maps
```{r}
ggplot() + 
    geom_sf(data=calgary, aes(fill=factor(ntile(allPredictions,5))), 
            colour=NA) +
    scale_fill_manual(values = c("#DEF5E5FF","#49C1ADFF","#357BA2FF","#3E356BFF","#0B0405FF"),
                      labels=as.character(quantile(calgary$allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                      name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  mapTheme +
   labs(title="Predicted Flooded Areas in Calgary")

ggplot() + 
  geom_sf(data=plvc, aes(fill=factor(ntile(plvc_allPredictions,5))), 
          colour=NA) +
  scale_fill_manual(values = c("#DEF5E5FF","#49C1ADFF","#357BA2FF","#3E356BFF","#0B0405FF"),
                    labels=as.character(quantile(plvc$plvc_allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                    name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  mapTheme +
  labs(title="Predicted Flooded Areas in Portland-Vancouver")
 

```

Calgary Predicted vs Observed
```{r}
 ggplot() + 
  geom_sf(data=calgary, aes(fill=factor(ntile(allPredictions,5))), colour=NA) +
  scale_fill_manual(values = c("#DEF5E5FF","#49C1ADFF","#357BA2FF","#3E356BFF","#0B0405FF"),
                    labels=as.character(quantile(calgary$allPredictions,
                                                 c(0.1,.2,.4,.6,.8),
                                                 na.rm=T)),
                    name="Predicted\nProbabilities(%)\n(Quintile\nBreaks)") +
  geom_sf(data=calgary  %>% 
               filter(MAJORITY == 1), 
               fill="gold",colour="NA") +
  mapTheme +
  labs(title="Observed and Predicted Flooded Areas",
       subtitle="Calgary; Observed flooded land in gold; Predicted flooded land in gradient ")
```

Calgary Confusion Metrics
```{r}
calgary %>%
  mutate(confResult=case_when(allPredictions < 50 & MAJORITY==0 ~ "True_Negative",
                              allPredictions >= 50 & MAJORITY==1 ~ "True_Positive",
                              allPredictions < 50 & MAJORITY==1 ~ "False_Negative",
                              allPredictions >= 50 & MAJORITY==0 ~ "False_Positive")) %>%
  ggplot()+
  geom_sf(aes(fill = confResult), color = "transparent")+
  scale_fill_manual(values = c("#DEF5E5FF","#49C1ADFF","#357BA2FF","#3E356BFF"),
                    name="Outcomes")+
  labs(title="Confusion Metrics For Calgary") +
  mapTheme
```





